A formal approach but practical approach in train regulation

### 2025-06-18 paper/main.tex
minor corrections

 **** Comparaison avec les travaux antérieurs ****

@@

We suppose that the train network can be described as follows. First, we consider given a set of blocks: $\blocks = \{ 1, 2, \ldots, n\}$. Blocks are oriented: there is a $\forward$ end and a $\backward$ one. Let $\directions = \{\forward, \backward\}$ be the two directions.  Next, we suppose given a set $\turnouts = \{ n+1, \ldots, m\}$, the identifiers  of trackworks (switches, crossings, double slip switches, and so on). Trackworks are characterized by an internal state: switches can be in direct or deviate position (next denoted '\direct' and '\deviate'),  a crossing has only one internal position, a double slip switch can be straight through or crossing. In the examples, to simplify the presentation, we use only switches, but our model is not restricted to them.  Let $\internalState$ denotes the set of internal states. A \emph{network configuration} is given by a function $S: \turnouts \to \internalState$. Let $\Sigma$ denotes the set of all network configurations. For two turnouts $\theta = \{9, 10\}$, the function $9 \mapsto \deviate, 10 \mapsto \direct$ states that $9$ is in deviate position while 10 is in direct one. The latter function is also described via an array  $[\deviate, \direct]$ giving states in trackwork order.

The network is then described as a triple $(\blocks, \turnouts, \sucblock)$ with $\sucblock: \blocks \times \directions \times \Sigma \to \blocks$. The successor function outputs the first block met according to the starting position, the direction and the network configuration.  From an abstract point of view, the topology of the network reduces to that data.

@@

Finally, aside of the tracks, at each end of the blocks, there is a signal whose state is \sigred (red) or \siggreen (green). In other words, the status of signals is described by a function $F: \blocks \times \directions \to \{ \sigred, \siggreen\}$. Given some signal status $F$, a block $pos$ and a direction $dir$, $F(pos,dir) = \sigred$ means a train in block $pos$ cannot leave it in direction $dir$. The signal status is entirely determined by the network configuration and the locking status due to the regulator. We will be back to it later.  Since we do not model train speed, we do not distinguish signals with speed limits from full-speed ones.

@@

To help the reader, in the sequel, by convention, we use $pid$ for blocks, $sid$ for trackworks identifiers (switch id) and $tid$ for train identifiers.

@@

\begin{grammar}[Regulator's rules][h][regulateur]
    \firstcasesubtil{stmt}{ev(tid, n) : (atom)*}{}
    \firstcasesubtil{atom}{\incr(pid)}{}
    \otherform{att(pid, v)}{}
    \otherform{\turn(sid, s)}{}
    \otherform{auth}{}
 \end{grammar}
with $tid \in \trains$, $n \in  \Nat$, $pid \in \blocks$, $sid \in \turnouts$, $s\in \internalState$.  Before we describe the intention of atoms and the other fields, a notation: given some function $f: X \to Y$, $x \in X$ and $y \in Y$, we denote by $f[x \leftarrow y]$, the function:
 $$f[x \leftarrow y] \triangleq \left\{  \begin{array}{ll} x &\mapsto y\\
x' \neq x \in X &\mapsto f(x')
\end{array}\right.$$

The token field of the regulator, denoted $T$, is a function  $T: \blocks \to \Nat$ associating a token level to each block.  The atom $\incr(pid)$ updates the token $pid$:  $T \cdot \incr(pid)  \triangleq T[pid \leftarrow T(pid)+1]$.

The waiting field is a function $W: \blocks \times \Nat \to \trains \cup \{0\}$. It states that block must be protected. More precisely, suppose that for some block $pid$, some train $tid$ tries to enter the block. If $W(pid,n) = tid$, then, the train shall wait the token is released. The value $W(pid, n) = 0$ means the train did not reach the current position $pid$.

>>>>

\todo{ Comparaison avec les travaux antérieurs }

@@

We suppose that the train network can be described as follows. First, we consider given a set of blocks: $\blocks = \{ 1, 2, \ldots, n\}$. Blocks are oriented: there is a $\forward$ (forward) end and a $\backward$ (backward) one. Let $\directions = \{\forward, \backward\}$ be the two directions.  Next, we suppose given a set $\turnouts = \{ n+1, \ldots, m\}$, the identifiers  of trackworks (switches, crossings, double slip switches, and so on). Trackworks are characterized by an internal state: switches can be in direct or deviate position (next denoted '\direct' and '\deviate'),  a crossing has only one internal position, a double slip switch can be straight through or crossing. In the examples, to simplify the presentation, we use only switches, but our model is not restricted to them.  Let $\internalState$ denotes the set of internal states. A \emph{network configuration} is given by a function $\sigma: \turnouts \to \internalState$. Let $\Sigma$ denotes the set of all network configurations. For two turnouts $\theta = \{9, 10\}$, the function $9 \mapsto \deviate, 10 \mapsto \direct$ states that $9$ is in deviate position while 10 is in direct one. The latter function is also described via an array  $[\deviate, \direct]$ giving states in trackwork order.

The network is then described as a triple $(\blocks, \turnouts, \sucblock)$ with $\sucblock: \blocks \times \directions \times \Sigma \to \blocks_\bot$. The successor function outputs the first block met according to the starting position, the direction and the network configuration.  The value $\sucblock(pid, d, \sigma) = \bot$ means that one reaches a dead-end. From an abstract point of view, the topology of the network reduces to that data.

@@

Finally, aside of the tracks, at each end of the blocks, there is a signal whose state is \sigred (red) or \siggreen (green). In other words, the status of signals is described by a function $F: \blocks \times \directions \to \{ \sigred, \siggreen\}$. Given some signal status $F$, a block $pid$ and a direction $d$, $F(pid,d) = \sigred$ means a train in block $pid$ cannot leave it in direction $d$. The signal status is entirely determined by the network configuration and the locking status due to the regulator. We will be back to it later.  Since we do not model train speed, we do not distinguish signals with speed limits from full-speed ones.

@@

To help the reader, by convention, we use $pid$ for blocks, $sid$ for trackworks identifiers (switch id) and $tid$ for train identifiers.

@@

\begin{grammar}[Regulator's rules][h][regulateur]
    \firstcasesubtil{stmt}{ev(tid, n) : (atom)*}{}
    \firstcasesubtil{atom}{\incr(pid)}{}
    \otherform{att(pid, v)}{}
    \otherform{\turn(sid, s)}{}
    \otherform{auth}{}
 \end{grammar}
with $tid \in \trains$, $n \in  \Nat$, $pid \in \blocks$, $sid \in \turnouts$, $s\in \internalState$.  
The token field of the regulator, denoted $T$, is a function  $T: \blocks \to \Nat$ associating a token level to each block.  The atom $\incr(pid)$ updates the token $pid$:  $T \cdot \incr(pid)  \triangleq T[pid \leftarrow T(pid)+1]$.

The waiting field is a function $W: \blocks \times \Nat \to \trains_0$. It states that block must be protected. More precisely, suppose that for some block $pid$, some train $tid$ tries to enter the block. If $W(pid,T(pid)) = tid$, then, the train shall wait the token is released. The value $W(pid, n) = 0$ means the train did not reach the current position $pid$.
